Baseline centering problem

### 2019-01-11 paper/tangotd.tex

\paragraph{Better variant?}
Everything works equally well if we just learn a kernel $Z(s_1,s_2)$ such
that
\begin{equation}
\transp{\1_{s_1}}\Delta^{-1}\psi\approx \E_{s_2\sim \rho} \,Z(s_1,s_2)
\psi(s_2)
\end{equation}
and this is still linear in $\psi$. This represents the inverse Laplace
operator by
\begin{equation}
Z=\sum_{s_1,s_2}\rho(s_2)  Z(s_1,s_2)\1_{s_1}\transp{\1_{s_2}}
\end{equation}
[TODO CHANGE NOTATION because currently the entries of $Z$ are not
$Z(s_1,s_2)$ but $Z_{s_1s_2}=\rho(s_2)Z(s_1,s_2)$ which is confusing...]
In particular the value function is
\begin{equation}
V(s_1)=\E_{s_2\sim \rho} \, Z(s_1,s_2) R(s_2)
\end{equation}
Downside: this representation of the value function does not allow for first
computing a representation of $R$ as $\E_{s_2} R(s_2)f(s_2)$ online, then taking a dot product with
$\green$. Here we would have to learn an auxiliary network to imitate
$\E_{s_2\sim \rho} Z(s_1,s_2) R(s_2)$ (but this is a supervised learning
problem; still, it looks suspiciously like $TD(1)$ except maybe for the
$O(1/(1-\gamma))$ variance). Alternatively, policy gradient can just directly be obtained
via
\begin{equation}
\partial \pi(a|s)\E_{s'|s,a} \E_{s_2\sim \rho} \, Z(s',s_2)R(s_2)
\end{equation}
(this works even if $\rho$ is \emph{not} the invariant distribution). We
can apply this each time a new reward is observed at some point $s_2$. We
can even set another arbitrary distribution on $s_1$: this policy
gradient will maximize
the reward obtained from a starting point from that distribution.

The left Bellman equation is
\begin{gather}
\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,P\Delta^{-1}\psi
\qquad \Longrightarrow
\\
\E_{s_2\sim \rho}
\,Z(s_1,s_2)\psi(s_2)
\approx \psi(s_1)+ \E_{s'\sim P(s'|s_1)}
\E_{s_2\sim \rho}\,
Z(s',s_2)\psi(s_2)
\end{gather}
for any $s_1$ and $\psi$.
The right Bellman equation becomes
\begin{gather}
\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,\Delta^{-1}P\psi
\qquad \Longrightarrow
\\
\E_{s_2\sim \rho}\,
Z(s_1,s_2)\psi(s_2)
\approx \psi(s_1)+ \E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}\,
Z(s_1,s_2)\psi(s'_2)
\end{gather}
These equations impose a kind of smoothness on $Z$ with respect to the
random walk $P$.

The Newton update/middle Bellman equation is
\begin{equation}
Z\gets (1+\eta)Z-\eta Z^2+\eta \gamma ZPZ
\end{equation}
which yields (after applying to $\1_{s_1}$ on the left and $\1_{s_2}/\rho(s_2)$
on the right)
\begin{multline}
Z(s_1,s_2)\gets (1+\eta)Z(s_1,s_2)-\eta \E_{s_3\sim \rho}\,
Z(s_1,s_3)Z(s_3,s_2)\,+\\\eta \gamma \E_{s_3\sim\rho,\, s'_3\sim
P(s'_3|s_3)}\,
Z(s_1,s_3)Z(s'_3,s_2)
\end{multline}
which is perfectly samplable. We can simulate the assignment
$Z(s_1,s_2)\gets...$ via gradient descent on the parameters of
$Z(s_1,s_2)$. \footnote{Should we adjust the size of the gradient descent
to get the right change of $Z(s_1,s_2)$? If we just use SGD, since
gradients may change from point to point, the speed may be different at
various states and we depart from Newton. This is clear when $Z$ is very
small: Newton grows exponentially while a standard gradient descent would
make small steps.} (Or via double sampling, which is not problematic
here!? but would converge to trivial solutions like $Z=0$ or
$Z=1/(1-\gamma)$, which are repulsive with Newton.)

Initialization: A first idea would be to start with paths from $s$ to
$s$, namely, initialize
$Z$ so that $Z\approx \Id$ at time
$0$, namely $Z(s,s)=1/\rho(s)$, although starting with $Z(s,s)=1$ should be
quite fine. (Small initializations of $Z$ are safer.) $Z>0$ (unless we
work with $\gamma=1$ and relative stuff), so the parameterization should
reflect that. Indeed the Newton iteration is unstable with negative
values.

Another initialization might be $Z(s_1,s_2)\approx \frac{1}{1-\gamma}$
for every pair $s_1,s_2$. Indeed, on-policy, $Z$ is the density with
respect to the invariant distribution $\mu$, and for $\gamma$ close to
$1$, $Z$ will be approximately $\frac{1}{1-\gamma}$ plus a smaller function (see
below). Note that $Z(s_1,s_2)=\frac{1}{1-\gamma}$ is an exact solution of
the middle Bellman equation (as is $Z=0$), but it is a repulsive fixed
point, I think, so initializing to $\frac{1}{1-\gamma}$ plus noise should
converge to the true solution. Initializing there should have the
advantage of starting close to reasonable values.

In the end this Newton equation alone might be sufficient to
encompass the left and right Bellman. It does not rely on test functions,
but the smoothing is implicit via the model used for $Z(s_1,s_2)$.

@@

Ideally, given the transition $s_3\to s'_3$, I'd like to sample $s_1$ and
$s_2$ proportionally to $Z$ (would cancel the $Z$ factors, and avoid
applying the equation to states that have nothing to do with
$s_3$—although the $Z$ factors are different for the three terms...). But
I don't know how to build an efficient sampler from $Z(s_3,\cdot)$.
Something like Metropolis? but this function $Z$ is not constant...

On-policy, sampling from $Z(\cdot,s_3)$ is easy: just sample one of the
recent states with geometric rates $\gamma$. This should divide the
right part of the equation by $Z(s_1,s_3)$ (up to some $\gamma$-dependent
factor). Thus we would sample two states $s_1,s_2$ along the trajectory,
sample a state $s_3$ in between on the trajectory from $s_1$, and (with
the right $\gamma$ factors) we would get a term with $Z(s_1,s_2)$, a term
with $Z(s_3,s_2)$ and a term with $Z(s'_3,s_2)$. This looks suspiciously
like a large-scale Bellman from $s_1$... Variance could be higher since
we don't use the model $Z$ and use sampling instead; plus, we lose any
generalization introduced by $Z$. To be tested. Write
carefully because $s'_3$ should be uncorrelated from $s_2$!! So not
possible with a single
trajectory?
On-policy, another
option is to sample triplets $(s_1,s_3,s_2)$ that are some time apart on
the trajectory, e.g., starting with $s_3$ uniform along the past
trajectory, and moving back for $s_1$ and forward for $s_2$ (with factors
$\gamma$). Looks a lot like eligibility traces?

On-policy, working with $\gamma\to 1$ is quite easy: indeed $Z(s_1,s_2)$ is
a density with respect to $\mu(s_2)$ so that the asymptotics when
$\gamma\to 1$ is $Z(s_1,s_2)=\frac{1}{1-\gamma}+Z_0(s_1,s_2)$, and we can
directly learn $Z_0$ by inserting $\frac{1}{1-\gamma}+Z_0$ in the middle
Bellman equation. This yields (omitting the expectations)
\begin{multline}
Z_0(s_1,s_2)\gets (1+\eta)Z_0(s_1,s_2)-\eta
Z_0(s_1,s_3)(Z_0(s_3,s_2)-\gamma Z_0(s'_3,s_2))\\-\eta
Z_0(s_1,s_3)+\frac{\eta}{1-\gamma}(\gamma Z_0(s'_3,s_2)-Z_0(s_3,s_2))
\end{multline}
TODO CHECK which we cannot directly apply with $\gamma=1$...
However ON-POLICY $s_3$ and
$s'_3$ have the same law, so in expectation the last term is $-\eta
\E_{s_3} Z_0(s_3,s_2)$. This yields the more symmetric
\begin{multline}
Z_0(s_1,s_2)\gets (1+\eta)Z_0(s_1,s_2)-\eta
Z_0(s_1,s_3)\left(Z_0(s_3,s_2)-\gamma Z_0(s'_3,s_2)\right)\\-\eta
Z_0(s_1,s_3)-\eta Z_0(s_3,s_2)
\end{multline}
on-policy, for any $\gamma\leq 1$.

@@

For discrete time but continuous state, the true $Z=\sum \gamma^t P^t$ is
equal to $\Id$ plus a continuous component; the $\Id$ is singular, so
learning $Z(s_1,s_2)$ using eg a neural network will be difficult around
the diagonal (and the initialization will be difficult). We may directly
write the dynamics on $Z-\Id$. (TODO not clear what this means in
continuous time: not clear if $\Delta^{-1}-\Id$ is actually small; still,
starting at $\Id$ is stable for Newton whatever $P$ is, so it should be
stable in the limit.)

@@

>>>>

\paragraph{Better variant?}
Everything works equally well if we just learn a kernel $z(s_1,s_2)$ such
that
\begin{equation}
\transp{\1_{s_1}}\Delta^{-1}\psi\approx \E_{s_2\sim \rho} \,z(s_1,s_2)
\psi(s_2)
\end{equation}
and this is still linear in $\psi$. This represents the inverse Laplace
operator by
\begin{equation}
\Delta^{-1}\approx Z=\sum_{s_1,s_2}\rho(s_2)  z(s_1,s_2)\1_{s_1}\transp{\1_{s_2}}
\end{equation}
In particular the value function is
\begin{equation}
V(s_1)=\E_{s_2\sim \rho} \, z(s_1,s_2) R(s_2)
\end{equation}
Downside: this representation of the value function does not allow for first
computing a representation of $R$ as $\E_{s_2} R(s_2)f(s_2)$ online, then taking a dot product with
$\green$. Here we would have to learn an auxiliary network to imitate
$\E_{s_2\sim \rho} z(s_1,s_2) R(s_2)$ (but this is a supervised learning
problem; still, it looks suspiciously like $TD(1)$ except maybe for the
$O(1/(1-\gamma))$ variance). Alternatively, policy gradient can just directly be obtained
via
\begin{equation}
\partial \pi(a|s)\E_{s'|s,a} \E_{s_2\sim \rho} \, z(s',s_2)R(s_2)
\end{equation}
(this works even if $\rho$ is \emph{not} the invariant distribution) or,
with a baseline,
\begin{equation}
\partial \pi(a|s)\E_{s'|s,a} \E_{s_2\sim \rho} \, (z(s',s_2)-z(s,s_2))R(s_2)
\end{equation}
which should have much better variance. We
can apply this each time a new reward is observed at some point $s_2$. We
can even set another arbitrary distribution on $s_1$: this policy
gradient will maximize
the reward obtained from a starting point from that distribution.

The left Bellman equation is
\begin{gather}
\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,P\Delta^{-1}\psi
\qquad \Longrightarrow
\\
\E_{s_2\sim \rho}
\,z(s_1,s_2)\psi(s_2)
\approx \psi(s_1)+ \E_{s'\sim P(s'|s_1)}
\E_{s_2\sim \rho}\,
z(s',s_2)\psi(s_2)
\end{gather}
for any $s_1$ and $\psi$.
The right Bellman equation becomes
\begin{gather}
\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,\Delta^{-1}P\psi
\qquad \Longrightarrow
\\
\E_{s_2\sim \rho}\,
z(s_1,s_2)\psi(s_2)
\approx \psi(s_1)+ \E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}\,
z(s_1,s_2)\psi(s'_2)
\end{gather}
These equations impose a kind of smoothness on $Z$ with respect to the
random walk $P$.

The Newton update/middle Bellman equation is
\begin{equation}
Z\gets (1+\eta)Z-\eta Z^2+\eta \gamma ZPZ
\end{equation}
which yields (after applying to $\1_{s_1}$ on the left and $\1_{s_2}/\rho(s_2)$
on the right)
\begin{multline}
z(s_1,s_2)\gets (1+\eta)z(s_1,s_2)-\eta \E_{s_3\sim \rho}\,
z(s_1,s_3)z(s_3,s_2)\,+\\\eta \gamma \E_{s_3\sim\rho,\, s'_3\sim
P(s'_3|s_3)}\,
z(s_1,s_3)z(s'_3,s_2)
\end{multline}
which is perfectly samplable. We can simulate the assignment
$z(s_1,s_2)\gets...$ via gradient descent on the parameters of
$z(s_1,s_2)$. \footnote{Should we adjust the size of the gradient descent
to get the right change of $z(s_1,s_2)$? If we just use SGD, since
gradients may change from point to point, the speed may be different at
various states and we depart from Newton. This is clear when $Z$ is very
small: Newton grows exponentially while a standard gradient descent would
make small steps.} (Or via double sampling, which is not problematic
here!? but would converge to trivial solutions like $Z=0$ or
$Z=1/(1-\gamma)$, which are repulsive with Newton.)

Initialization: A first idea would be to start with paths from $s$ to
$s$, namely, initialize
$Z$ so that $Z\approx \Id$ at time
$0$, namely $z(s,s)=1/\rho(s)$, although starting with $z(s,s)=1$ should be
quite fine. (Small initializations of $Z$ are safer.) $Z>0$ (unless we
work with $\gamma=1$ and relative stuff), so the parameterization should
reflect that. Indeed the Newton iteration is unstable with negative
values.

Another initialization might be $z(s_1,s_2)\approx \frac{1}{1-\gamma}$
for every pair $s_1,s_2$. Indeed, on-policy, $z$ is the density with
respect to the invariant distribution $\mu$, and for $\gamma$ close to
$1$, $z$ will be approximately $\frac{1}{1-\gamma}$ plus a smaller function (see
below). Note that $z(s_1,s_2)=\frac{1}{1-\gamma}$ is an exact solution of
the middle Bellman equation (as is $z=0$), but it is a repulsive fixed
point, I think, so initializing to $\frac{1}{1-\gamma}$ plus noise should
converge to the true solution [empirically I find this diverges]. Initializing there should have the
advantage of starting close to reasonable values.

In the end this Newton equation alone might be sufficient to
encompass the left and right Bellman. It does not rely on test functions,
but the smoothing is implicit via the model used for $z(s_1,s_2)$.

@@

Ideally, given the transition $s_3\to s'_3$, I'd like to sample $s_1$ and
$s_2$ proportionally to $Z$ (would cancel the $Z$ factors, and avoid
applying the equation to states that have nothing to do with
$s_3$—although the $Z$ factors are different for the three terms...). But
I don't know how to build an efficient sampler from $z(s_3,\cdot)$.
Something like Metropolis? but this function $Z$ is not constant...

On-policy, sampling from $z(\cdot,s_3)$ is easy: just sample one of the
recent states with geometric rates $\gamma$. This should divide the
right part of the equation by $z(s_1,s_3)$ (up to some $\gamma$-dependent
factor). Thus we would sample two states $s_1,s_2$ along the trajectory,
sample a state $s_3$ in between on the trajectory from $s_1$, and (with
the right $\gamma$ factors) we would get a term with $z(s_1,s_2)$, a term
with $z(s_3,s_2)$ and a term with $z(s'_3,s_2)$. This looks suspiciously
like a large-scale Bellman from $s_1$... Variance could be higher since
we don't use the model $Z$ and use sampling instead; plus, we lose any
generalization introduced by $Z$. To be tested. Write
carefully because $s'_3$ should be uncorrelated from $s_2$!! So not
possible with a single
trajectory?
On-policy, another
option is to sample triplets $(s_1,s_3,s_2)$ that are some time apart on
the trajectory, e.g., starting with $s_3$ uniform along the past
trajectory, and moving back for $s_1$ and forward for $s_2$ (with factors
$\gamma$). Looks a lot like eligibility traces?

On-policy, working with $\gamma\to 1$ is quite easy: indeed $z(s_1,s_2)$ is
a density with respect to $\mu(s_2)$ so that the asymptotics when
$\gamma\to 1$ is $z(s_1,s_2)=\frac{1}{1-\gamma}+z_0(s_1,s_2)$, and we can
directly learn $z_0$ by inserting $\frac{1}{1-\gamma}+z_0$ in the middle
Bellman equation. This yields (omitting the expectations)
\begin{multline}
z_0(s_1,s_2)\gets (1+\eta)z_0(s_1,s_2)-\eta
z_0(s_1,s_3)(z_0(s_3,s_2)-\gamma z_0(s'_3,s_2))\\-\eta
z_0(s_1,s_3)+\frac{\eta}{1-\gamma}(\gamma z_0(s'_3,s_2)-z_0(s_3,s_2))
\end{multline}
TODO CHECK which we cannot directly apply with $\gamma=1$...
However ON-POLICY $s_3$ and
$s'_3$ have the same law, so in expectation the last term is $-\eta
\E_{s_3} z_0(s_3,s_2)$. This yields the more symmetric
\begin{multline}
z_0(s_1,s_2)\gets (1+\eta)z_0(s_1,s_2)-\eta
z_0(s_1,s_3)\left(z_0(s_3,s_2)-\gamma z_0(s'_3,s_2)\right)\\-\eta
z_0(s_1,s_3)-\eta z_0(s_3,s_2)
\end{multline}
on-policy, for any $\gamma\leq 1$.

@@

For discrete time but continuous state, the true $Z=\sum \gamma^t P^t$ is
equal to $\Id$ plus a continuous component; the $\Id$ is singular, so
learning $z(s_1,s_2)$ using eg a neural network will be difficult around
the diagonal (and the initialization will be difficult). We may directly
write the dynamics on $Z-\Id$. (TODO not clear what this means in
continuous time: not clear if $\Delta^{-1}-\Id$ is actually small; still,
starting at $\Id$ is stable for Newton whatever $P$ is, so it should be
stable in the limit.)

@@

Normalization of $Z$. Since $Z=\sum \gamma^t P^t$, we have that
$(1-\gamma)Z$ is a stochastic matrix (thanks Yann). This might help with
normalization/detecting divergence. This translates as $\E_{s_2\sim \rho}
z(s_1,s_2)=\frac{1}{1-\gamma}$ which is testable/enforceable. However, I
fear it might make the fixed point $z(s_1,s_2)=\frac{1}{1-\gamma}$
attractive instead of repulsive, and it might also perturb initialization
(initializing $Z=\Id$ does not satisfy this normalization, so enforcing
it too early is basically like initializing to
$Z=\frac{1}{1-\gamma}\Id$ which empirically diverges).
